Application of Equal Local Levels to Improve Q-Q Plot Testing Bands with \proglang{R} Package \pkg{qqconf}

### 2021-04-26 article/manuscript.Rnw
pushed article updates

\title{Efficient Calculation of Improved Simultaneous Testing Bands for QQplots with \proglang{R} Package \pkg{qqconf}}
\Plaintitle{Efficient Calculation of Improved Simultaneous Testing Bands for QQplots with R Package qqconf}
\Shorttitle{Efficient Calculation of Simultaneous Testing Bands for QQplots with \pkg{qqconf}}
%\author{Eric Weine and Mary Sara McPeek}
\Abstract{
  Quantile-Quantile plots (QQplots) are often difficult to interpret because it is unclear how large the deviation from the theoretical distribution must be to indicate a lack of fit. Current packages and algorithms towards this end either do not ensure a robust Type I error rate, are too slow, or are under-powered to deviations in the tails of the distribution. In this paper, we present an efficient algorithm and accompanying \proglang{R} Package \pkg{qqconf} that computes simultaneous testing bands for QQplots. These bands have a variety of desirable properties, including being fast to compute and being equally sensitive to deviations in all parts of the null distribution, including the tails.
}

\Keywords{QQplots, Equal Local Levels, Kolmolgorov-Smirnov, GWAS, Multiple Testing}

@@

Quantile-Quantile plots (QQplots) are a common statistical tool used to judge if a sample comes from a specified distribution \citep{wilk_gnanadesikan_1968}. Despite their ubiquity, they are often difficult to interpret because it is challenging to determine if the magnitude of the deviation from the specified distribution is large enough to indicate a lack of fit as opposed to sampling variability. To make this determination, it is useful to put testing bands on a QQplot.

@@

A number of methods have been created towards this end. Naively, one could put pointwise testing bands onto QQplots, which conduct $\alpha$ level tests on each order statistic of the sample. This method is clearly insufficient, as it ignores the multiple testing problem and thus provides no guarantee on the Type I error rate of testing the global null hypothesis that all of the data come from the specified null distribution. To appropriately deal with this multiple testing problem, the Komolgorov-Smirnov (KS) statistic is commonly used to create joint testing bands \citep{kolmogoroff1941confidence, smirnov1944approximate}. While this method ensures the correct Type I error by using the distribution of the KS statistic to create joint testing bands, this test suffers from very low power under a variety of reasonable alternatives since it has low sensitivity to deviation in the tails of the null distribution \citep{aldor2013power}. A more promising approach, originally introduced by Berk and Jones and later developed by Gontscharuk et al., conducts a test with Type I error rate $\eta$ on each order statistic of the sample such that the Type I error rate of the global test is some desired level $\alpha$ \citep{berk1979goodness, gontscharuk2016goodness}. Because this test, which will be referred to as equal local levels (ELL), conducts an $\eta$ level test on each order statistic, it is agnostic to any alternative distribution. This makes ELL a sensible choice for a standard test to use in creating testing bands on QQplots, as it does not bias the user towards deviations of the sample in any particular part of the null distribution.

@@

Currently, there is no available software that allows a user to quickly create plausible and robust testing bands for QQplots on real data. While the base-\proglang{R} package \pkg{stats} provides functionality for creating a QQplot to compare a sample against the normal distribution, it doesn't give the user the ability to create QQplots for other distributions or to put testing bands on those plots. The package \pkg{qqplotr} provides a number of helpful additions to the base-\proglang{R} functionality, including the ability to create QQplots for a variety of reference distributions and a robust method for testing bands on the normal distribution \citep{qqplotr}. However, this package does not provide adequate testing bands for non-normal reference distributions and is not able to efficiently create testing bands for large sample sizes.

@@

In what follows, we introduce the \proglang{R} package \pkg{qqconf}. With the aid of a novel recursive calculation developed below, \pkg{qqconf} can quickly provide ELL testing bands for comparing even very large samples to any reference distribution with a CDF implemented. In addition to these testing bands, \pkg{qqconf} provides a variety of plotting functionalities that allow the user to easily visualize where any deviation of the sample from the null distribution may occur. In section \hyperref[sec:methods]{2}, we introduce the methods required for the efficient recursive calculation mentioned above. In section \hyperref[sec:methods]{3}, we discuss the behavior of ELL bands in large sample sizes and an approximation to these bands that makes them very easy to calculate for even very large samples ($n > 5 \cdot 10^{6}$). In section \hyperref[sec:examples]{4}, we introduce \pkg{qqconf} and demonstrate its functionality on a few multiple testing scenarios including including Type I error calibration and genetic studies. Finally, we conclude in section \hyperref[sec:discussion]{5} with a discussion of limitations the of ELL and \pkg{qqconf} as well as further directions for creating testing bands on QQplots.

@@

One of the main advantages of the local levels method compared to other global testing approaches, is that it can easily be used to put testing bands onto QQplots by simply graphing each $(h_{i}, g_{i})$ interval. This allows us to examine how a dataset might deviate from some null distribution much better than simply applying a test that yields a binary conclusion. Below, we present a few examples where a QQplot is useful, and where the local levels test seems ideal for assessing deviation from a global null hypothesis.

@@

Figure 3 shows the resulting QQplots for scenarios 1 (in blue) and 2 (in red), where the 45$^{\mbox{\footnotesize o}}$ line is shown as well as the testing bands obtained from the equal local levels procedure for testing, at global level .05, the null hypothesis that $p_1, \ldots, p_n$ have the same distribution as $n$ i.i.d.\ draws from Unif(0,1).

@@

\begin{figure}
\begin{center}
\includegraphics{manuscript-chisq_qq}
\end{center}
\caption{QQplot for Scenario 1 and Scenario 2 With Standard Axes.}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{manuscript-chisq_qq_log}
\end{center}
\caption{QQplot for Scenario 1 and Scenario 2 With Axes On Log Scale.}
\end{figure}

@@

The local levels test using an exact recursive calculation allows for the creation of relatively fast and alternative-agnostic simultaneous testing bands for QQplots. Of course, our method is limited by the assumptions that (1) all of the tests are independent and that (2) all of the parameters of the null distribution are known. Currently, we are not aware of methods to handle the dependent case that allow for bands on QQplots. To handle the second issue, Aldor-Noiman has adapted their method such that it will yield exact results when some parameters of the null distribution are unknown. However, if the dataset is very large, Aldor-Noiman's method is prohibitively slow. In this case, we suggest that our method be used because it is much faster, and with a sufficiently large dataset the uncertainty in the parameters will become negligible under standard regularity conditions. Methods to handle parameter uncertainty with an exact calculation have been discussed in the context of the normal distribution \citep{rosenkrantz2000confidence}, but a general method towards this end has not been developed, and would be an important next step.

>>>>

\title{Application of Equal Local Levels to Improve Q-Q Plot Testing Bands with \proglang{R} Package \pkg{qqconf}}
\Plaintitle{Application of Equal Local Levels to Improve Q-Q Plot Testing Bands with R Package qqconf}
\Shorttitle{Equal Local Levels Q-Q Plot Testing Bands with \pkg{qqconf}}
\Abstract{
  Quantile-Quantile plots (Q-Q plots) are often difficult to interpret because it is unclear how large the deviation from the theoretical distribution must be to indicate a lack of fit. Current packages and algorithms towards this end either do not ensure a robust Type I error rate, are too slow, or are under-powered to deviations in the tails of the distribution. In this paper, we present an efficient algorithm and accompanying \proglang{R} Package \pkg{qqconf} that computes simultaneous testing bands for Q-Q plots. These bands have a variety of desirable properties, including being fast to compute and being equally sensitive to deviations in all parts of the null distribution, including the tails.
}

\Keywords{Q-Q plots, Equal Local Levels, Kolmolgorov-Smirnov, GWAS, Multiple Testing}

@@

Quantile-Quantile plots (Q-Q plots) are a common statistical tool used to judge if a sample comes from a specified distribution \citep{wilk_gnanadesikan_1968}. Despite their ubiquity, they are often difficult to interpret because it is challenging to determine if the magnitude of the deviation from the specified distribution is large enough to indicate a lack of fit as opposed to sampling variability. To make this determination, it is useful to put testing bands on a Q-Q plot.

@@

A number of methods have been created towards this end. Naively, one could put pointwise testing bands onto Q-Q plots, which conduct $\alpha$ level tests on each order statistic of the sample. This method is clearly insufficient, as it ignores the multiple testing problem and thus provides no guarantee on the Type I error rate of testing the global null hypothesis that all of the data come from the specified null distribution. To appropriately deal with this multiple testing problem, the Komolgorov-Smirnov (KS) statistic is commonly used to create joint testing bands \citep{kolmogoroff1941confidence, smirnov1944approximate}. While this method ensures the correct Type I error by using the distribution of the KS statistic to create joint testing bands, this test suffers from very low power under a variety of reasonable alternatives since it has low sensitivity to deviation in the tails of the null distribution \citep{aldor2013power, berk1979goodness}. A more promising approach would be to apply the equal local levels method (ELL) to create testing bands. This method, originally introduced by Berk and Jones and later developed by Gontscharuk et al. as an alternative testing procedure to higher-criticism and KS, conducts a test with Type I error rate $\eta$ on each order statistic of the sample such that the Type I error rate of the global test is some desired level $\alpha$ \citep{berk1979goodness, gontscharuk2015intermediates, gontscharuk2016goodness, gontscharuk2017asymptotics}. Because this test conducts an $\eta$ level test on each order statistic, it can be used to create testing bands that are agnostic to any alternative distribution. While the ELL method has never previously been specifically applied to creating testing bands on Q-Q plots, we suggest it is a sensible choice for a standard test in this use case, as it does not bias the user towards deviations of the sample in any particular part of the null distribution.

@@

Currently, there is no available software that allows a user to quickly create plausible and robust testing bands for Q-Q plots on real data. While the base-\proglang{R} package \pkg{stats} provides functionality for creating a Q-Q plot to compare a sample against the normal distribution, it doesn't give the user the ability to create Q-Q plots for other distributions or to put testing bands on those plots. The package \pkg{qqplotr} provides a number of helpful additions to the base-\proglang{R} functionality, including the ability to create Q-Q plots for a variety of reference distributions and a robust method for testing bands on the normal distribution \citep{qqplotr}. Specifically, for the normal distribution \pkg{qqplotr} implements a simulation based method called tail sensitive (TS) testing bands, originally developed by Alder-Noiman \citep{aldor2013power}.  However, this package does not provide adequate testing bands for non-normal reference distributions and is not able to efficiently create testing bands for large sample sizes because it uses simulation.

@@

In what follows, we introduce the \proglang{R} package \pkg{qqconf}. With the aid of a novel recursive calculation developed below, \pkg{qqconf} can quickly provide ELL testing bands for comparing even very large samples to any reference distribution with a CDF implemented. In addition to these testing bands, \pkg{qqconf} provides a variety of plotting functionalities that allow the user to easily visualize where any deviation of the sample from the null distribution may occur. In section \hyperref[sec:methods]{2}, we introduce the methods required for the efficient recursive calculation mentioned above. In section \hyperref[sec:methods]{3}, we discuss the behavior of ELL bands in large sample sizes and an approximation to these bands that makes them very easy to calculate for even very large samples ($n > 5 \cdot 10^{6}$). In section \hyperref[sec:examples]{4}, we introduce \pkg{qqconf} and demonstrate its functionality on a few multiple testing scenarios including including Type I error calibration and genetic studies. Finally, we conclude in section \hyperref[sec:discussion]{5} with a discussion of limitations the of ELL and \pkg{qqconf} as well as further directions for creating testing bands on Q-Q plots.

@@

One of the main advantages of the local levels method compared to other global testing approaches, is that it can easily be used to put testing bands onto Q-Q plots by simply graphing each $(h_{i}, g_{i})$ interval. This allows us to examine how a dataset might deviate from some null distribution much better than simply applying a test that yields a binary conclusion. Below, we present a few examples where a Q-Q plot is useful, and where the local levels test seems ideal for assessing deviation from a global null hypothesis.

@@

Figure 3 shows the resulting Q-Q plots for scenarios 1 (in blue) and 2 (in red), where the 45$^{\mbox{\footnotesize o}}$ line is shown as well as the testing bands obtained from the equal local levels procedure for testing, at global level .05, the null hypothesis that $p_1, \ldots, p_n$ have the same distribution as $n$ i.i.d.\ draws from Unif(0,1).

@@

\begin{figure}
\begin{center}
\includegraphics{manuscript-chisq_qq}
\end{center}
\caption{Q-Q plot for Scenario 1 and Scenario 2 With Standard Axes.}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics{manuscript-chisq_qq_log}
\end{center}
\caption{Q-Q plot for Scenario 1 and Scenario 2 With Axes On Log Scale.}
\end{figure}

@@

The local levels test using an exact recursive calculation allows for the creation of relatively fast and alternative-agnostic simultaneous testing bands for Q-Q plots. Of course, our method is limited by the assumptions that (1) all of the tests are independent and that (2) all of the parameters of the null distribution are known. Currently, we are not aware of methods to handle the dependent case that allow for bands on Q-Q plots. To handle the second issue, Aldor-Noiman has adapted their method such that it will yield exact results when some parameters of the null distribution are unknown. However, if the dataset is very large, Aldor-Noiman's method is prohibitively slow. In this case, we suggest that our method be used because it is much faster, and with a sufficiently large dataset the uncertainty in the parameters will become negligible under standard regularity conditions. Methods to handle parameter uncertainty with an exact calculation have been discussed in the context of the normal distribution \citep{rosenkrantz2000confidence}, but a general method towards this end has not been developed, and would be an important next step.

### 2021-04-26 article/manuscript.tex
pushed article updates

A number of methods have been created towards this end. Naively, one could put pointwise testing bands onto QQplots, which conduct $\alpha$ level tests on each order statistic of the sample. This method is clearly insufficient, as it ignores the multiple testing problem and thus provides no guarantee on the Type I error rate of testing the global null hypothesis that all of the data come from the specified null distribution. To appropriately deal with this multiple testing problem, the Komolgorov-Smirnov (KS) statistic is commonly used to create joint testing bands \citep{kolmogoroff1941confidence, smirnov1944approximate}. While this method ensures the correct Type I error by using the distribution of the KS statistic to create joint testing bands, this test suffers from very low power under a variety of reasonable alternatives since it has low sensitivity to deviation in the tails of the null distribution \citep{aldor2013power}. A more promising approach, originally introduced by Berk and Jones and later developed by Gontscharuk et al., conducts a test with Type I error rate $\eta$ on each order statistic of the sample such that the Type I error rate of the global test is some desired level $\alpha$ \citep{berk1979goodness, gontscharuk2016goodness}. Because this test, which will be referred to as equal local levels (ELL), conducts an $\eta$ level test on each order statistic, it is agnostic to any alternative distribution. This makes ELL a sensible choice for a standard test to use in creating testing bands on QQplots, as it does not bias the user towards deviations of the sample in any particular part of the null distribution.

@@

Currently, there is no available software that allows a user to quickly create plausible and robust testing bands for QQplots on real data. While the base-\proglang{R} package \pkg{stats} provides functionality for creating a QQplot to compare a sample against the normal distribution, it doesn't give the user the ability to create QQplots for other distributions or to put testing bands on those plots. The package \pkg{qqplotr} provides a number of helpful additions to the base-\proglang{R} functionality, including the ability to create QQplots for a variety of reference distributions and a robust method for testing bands on the normal distribution \citep{qqplotr}. However, this package does not provide adequate testing bands for non-normal reference distributions and is not able to efficiently create testing bands for large sample sizes.

@@

In what follows, we introduce the \proglang{R} package \pkg{qqconf}. With the aid of a novel recursive calculation developed below, \pkg{qqconf} can quickly provide ELL testing bands for comparing even very large samples to any reference distribution with a CDF implemented. In addition to these testing bands, \pkg{qqconf} provides a variety of plotting functionalities that allow the user to easily visualize where any deviation of the sample from the null distribution may occur. In section \hyperref[sec:methods]{2}, we introduce the methods required for the efficient recursive calculation mentioned above. In section \hyperref[sec:methods]{3}, we discuss the behavior of ELL bands in large sample sizes and an approximation to these bands that makes them very easy to calculate for even very large samples ($n > 5 \cdot 10^{6}$). In section \hyperref[sec:examples]{4}, we introduce \pkg{qqconf} and demonstrate its functionality on a few multiple testing scenarios including including Type I error calibration and genetic studies. Finally, we conclude in section \hyperref[sec:discussion]{5} with a discussion of limitations the of ELL and \pkg{qqconf} as well as further directions for creating testing bands on QQplots.

>>>>

A number of methods have been created towards this end. Naively, one could put pointwise testing bands onto Q-Q plots, which conduct $\alpha$ level tests on each order statistic of the sample. This method is clearly insufficient, as it ignores the multiple testing problem and thus provides no guarantee on the Type I error rate of testing the global null hypothesis that all of the data come from the specified null distribution. To appropriately deal with this multiple testing problem, the Komolgorov-Smirnov (KS) statistic is commonly used to create joint testing bands \citep{kolmogoroff1941confidence, smirnov1944approximate}. While this method ensures the correct Type I error by using the distribution of the KS statistic to create joint testing bands, this test suffers from very low power under a variety of reasonable alternatives since it has low sensitivity to deviation in the tails of the null distribution \citep{aldor2013power, berk1979goodness}. A more promising approach would be to apply the equal local levels method (ELL) to create testing bands. This method, originally introduced by Berk and Jones and later developed by Gontscharuk et al. as an alternative testing procedure to higher-criticism and KS, conducts a test with Type I error rate $\eta$ on each order statistic of the sample such that the Type I error rate of the global test is some desired level $\alpha$ \citep{berk1979goodness, gontscharuk2015intermediates, gontscharuk2016goodness, gontscharuk2017asymptotics}. Because this test conducts an $\eta$ level test on each order statistic, it can be used to create testing bands that are agnostic to any alternative distribution. While the ELL method has never previously been specifically applied to creating testing bands on Q-Q plots, we suggest it is a sensible choice for a standard test to use in creating testing bands on Q-Q plots, as it does not bias the user towards deviations of the sample in any particular part of the null distribution.

@@

Currently, there is no available software that allows a user to quickly create plausible and robust testing bands for Q-Q plots on real data. While the base-\proglang{R} package \pkg{stats} provides functionality for creating a Q-Q plot to compare a sample against the normal distribution, it doesn't give the user the ability to create Q-Q plots for other distributions or to put testing bands on those plots. The package \pkg{qqplotr} provides a number of helpful additions to the base-\proglang{R} functionality, including the ability to create Q-Q plots for a variety of reference distributions and a robust method for testing bands on the normal distribution \citep{qqplotr}. However, this package does not provide adequate testing bands for non-normal reference distributions and is not able to efficiently create testing bands for large sample sizes.

@@

%Add a line about Alder Noiman here where you mention tail sensitive and qqplotr
In what follows, we introduce the \proglang{R} package \pkg{qqconf}. With the aid of a novel recursive calculation developed below, \pkg{qqconf} can quickly provide ELL testing bands for comparing even very large samples to any reference distribution with a CDF implemented. In addition to these testing bands, \pkg{qqconf} provides a variety of plotting functionalities that allow the user to easily visualize where any deviation of the sample from the null distribution may occur. In section \hyperref[sec:methods]{2}, we introduce the methods required for the efficient recursive calculation mentioned above. In section \hyperref[sec:methods]{3}, we discuss the behavior of ELL bands in large sample sizes and an approximation to these bands that makes them very easy to calculate for even very large samples ($n > 5 \cdot 10^{6}$). In section \hyperref[sec:examples]{4}, we introduce \pkg{qqconf} and demonstrate its functionality on a few multiple testing scenarios including including Type I error calibration and genetic studies. Finally, we conclude in section \hyperref[sec:discussion]{5} with a discussion of limitations the of ELL and \pkg{qqconf} as well as further directions for creating testing bands on Q-Q plots.
